Left-linear rules in adhesive categories

### 2024-02-14 icalp/fusione.tex
minori

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ `and  $\dder{D}_j = \dder{E}_j$ for each $j \not \in \{i,i+1\}$. We write
  $\der{D} \shift{\transp{i}} \der{E}$.

@@

\subsection{well-switching rewriting systems}
\label{subsec:verytame}

@@

\begin{definition}[Graph signature]
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph where $s,t\colon E\rightrightarrows V$.
  We define the \emph{depth} of a vertex $v\in V$ as
  \[
    \dph(v)=\begin{cases}
      0 & \text{there is no $e\in E$ such that $t(e)=v$}\\
      m+1 &\text{where } m=\min\{\dph(s(e)) \mid e\in E \text{ such that } t(e)=v
      \} 
    \end{cases}
  \]
\end{definition}

@@

\todo{cite something here? graph signatures?}

@@

Left-linear rewriting systems on $\cat{eGraphs}$, i.e. in which the right half of a rule is mono but not regular in general, 
are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one at the receiver,
but put instead in the same equivalence class, thus better tracing the causal dependencies among processes reductions.
\end{example}

@@

\begin{example}
  The result above fails when considering left-linear
  derivations. This is readily seen by considering derivations
  $\der{D}$ in Fig.~\ref{fi:derD} and $\der{D}'$ in
  Fig.~\ref{fi:derD1}. We already observed that
  $\der{D} \not\equiv_a \der{D}'$\todo{P: osservato?}
  %
  Still, it is easy to see that they are related by a consistent
  permutation, which is the identity on productions. In fact, the colimit of
  both derivations is a single node with a self loop, i.e., the
  terminal object in $\cat{Graph}$, hence all commutation requirements
  are trivially satisfied.
\end{example}

@@

This result clearly fails when considering left-linear
derivations. This is readily seen by considering derivations $\der{D}$
in Fig.~\ref{fi:derD} and $\der{D}'$ in Fig.~\ref{fi:derD1}. We
already observed that $\der{D} \not\equiv_a \der{D}'$\todo{P:
  osservato?}
% 
Still, it is easy to see that they are related by a consistent
permutation, which is the identity on productions. The colimit of
both derivations is a single node with a self loop, i.e. the
terminal object in $\cat{Graph}$, hence all commutation requirements
are trivially satisfied.

>>>>

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ and  $\dder{D}_j = \dder{E}_j$ for each $j \not \in \{i,i+1\}$. We write
  $\der{D} \shift{\transp{i}} \der{E}$.

@@

\subsection{Well-switching rewriting systems}
\label{subsec:verytame}

@@

\begin{definition}
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph where $s,t\colon E\rightrightarrows V$.
  We define the \emph{depth} of a vertex $v\in V$ as
  \[
    \dph(v)=\begin{cases}
      0 & \text{there is no $e\in E$ such that $t(e)=v$}\\
      m+1 &\text{where } m=\min\{\dph(s(e)) \mid e\in E \text{ such that } t(e)=v
      \} 
    \end{cases}
  \]
\end{definition}

@@

@@

Left-linear rewriting systems on $\cat{EGraphs}$, i.e. in which the right half of a rule is mono but not regular in general, 
are well-switching. They have been used in~\cite{BaldanGM06} for the graphical implementation of nominal calculi, where,
differently from~\cite{Gad07}, as a result of name passing the received name is not merged with a local one at the receiver,
but put instead in the same equivalence class, thus better tracing the causal dependencies among processes reductions.
\end{example}

@@

\begin{example}
  The result above fails when considering left-linear
  derivations. This is readily seen by considering derivations
  $\der{D}$ in Fig.~\ref{fi:derD} and $\der{D}'$ in
  Fig.~\ref{fi:derD1}. We already observed that
  $\der{D}$ and $\der{D}'$ are not switch equivalent.
  %
  Still, it is easy to see that they are related by a consistent
  permutation, which is the identity on productions. In fact, the colimit of
  both derivations is a single node with a self loop, i.e., the
  terminal object in $\cat{Graph}$, hence all commutation requirements
  are trivially satisfied.
\end{example}

@@

This result clearly fails when considering left-linear
derivations. This is readily seen by considering derivations $\der{D}$
in Fig.~\ref{fi:derD} and $\der{D}'$ in Fig.~\ref{fi:derD1}. We
already observed that $\der{D}$ and $\der{D}'$ are not switch equivalent.
% 
Still, it is easy to see that they are related by a consistent
permutation, which is the identity on productions. The colimit of
both derivations is a single node with a self loop, i.e. the
terminal object in $\cat{Graph}$, hence all commutation requirements
are trivially satisfied.
